---
title: "Analysis of the maker-generated OGS gene set using kallisto and sleuth"
output: html_notebook
---

# Set up

Load required libs

```{r}
library(sleuth) #also loads dplyr
library(cowplot)
library(stringr)
```

Next set up the "auxilliary"" table with sample information and paths to data

```{r sample.info}
kallisto.path = "../kallisto/ogs"
sample.info <- data.frame(sample = list.files(kallisto.path, pattern = "*_rep*"), stringsAsFactors = F)
sample.info$path <- paste(kallisto.path, sample.info$sample, sep = "/")
tmp <- str_extract(sample.info$sample, pattern = "^.")
sample.info$treatment <- factor(case_when(tmp == "C" ~ "Control",
                                          tmp == "E" ~ "Esfenvalerate",
                                          tmp == "T" ~ "Thiamethoxam"),
                                levels = c("Control", "Esfenvalerate", "Thiamethoxam"))
rm(tmp)
sample.info$time <- as.integer(str_sub(sample.info$sample, start = 2, end = 3))
sample.info <- sample.info[,c(1,3,4,2)]

sample.info
```


We will also want to load the BLAST2GO and Interpro scan results for functional information about any differentially expressed genes.

```{r b2g.annotations}
b2g.annotations <- read.table("../genome/SBA_annot/blast2go.annot",
                              col.names = c("protein_id", "GO.term", "name"),
                              stringsAsFactors = F,sep = '\t')
head(b2g.annotations)
```


Note that the downloaded file of IPS results does not include a header line. Header info comes from ![https://github.com/ebi-pf-team/interproscan/wiki/OutputFormats]

```{r ips.annotations}

ips.annotations <- read.table("../genome/SBA_annot/interproscan.tsv", 
                              sep = '\t', 
                              stringsAsFactors = F, 
                              fill = T)

names(ips.annotations) <- c("protein_id",
                            "seq.md5",
                            "seq.length",
                            "analysis",
                            "accession",
                            "description",
                            "start",
                            "stop",
                            "score",
                            "status",
                            "date",
                            "interpro.acc",
                            "interpro.desc",
                            "GO",
                            "pathways")
```


# Quality control

A nice thing about Sleuth is that is allows us to spot any outlier samples that might be throwing off our analysis. A first step is to look at a PCA plot.

```{r qc_PCA}
all.sleuth <- sleuth_prep(sample.info,
                          extra_bootstrap_summary = T)

plot_pca(all.sleuth, color_by = 'treatment', text_labels = T)
```
So, it looks like we have some potential outlier issues here, most notably with C12_rep4, but T12_rep25 and C12_rep3 are also a little suspect.

I have been wondering if there is something funny going on with the 12 hour samples. Maybe we would do better to start by looking at the 24 hour samples.


# Esfenvalerate at 12 hours

We have already looked that this with DESeq2, so it is good to make the comparison.

Subset the sample info

```{r off.e12.info, eval=FALSE, include=FALSE}
e12.info <- filter(sample.info, 
                   treatment %in% c("Control", "Esfenvalerate"), 
                   time == 12)

e12.info
```

Make a sleuth object

```{r off.e12.sleuth, eval=FALSE, include=FALSE}
e12.sleuth <- sleuth_prep(e12.info,
                          extra_bootstrap_summary = T)
```

Next fit the full model, with treatments and the reduced model with no treatments and perform liklihood ratio test.

```{r off.lrt_e12, eval=FALSE, include=FALSE}
e12.sleuth <- sleuth_fit(e12.sleuth,
                         formula = ~treatment,
                         fit_name = "full")

e12.sleuth <- sleuth_fit(e12.sleuth,
                         formula = ~1,
                         fit_name = "reduced")

e12.sleuth <- sleuth_lrt(e12.sleuth, "reduced", "full")

models(e12.sleuth)
tests(e12.sleuth)
```

Extract the LRT test results

```{r off.e12.results, eval=FALSE, include=FALSE}

e12.results <- sleuth_results(e12.sleuth, test = 'reduced:full', test_type = 'lrt', show_all = FALSE)
e12.results

```

Get the results that are significant at FDR <= 0.1

```{r off.e12.sig.results, eval=FALSE, include=FALSE}
e12.sig.results <- filter(e12.results,
                          qval <= 0.1)


```

How many sig results do we have?

```{r eval=FALSE, include=FALSE}
length(e12.sig.results$target_id)
```

This seems like a lot given there are 42274 total transcripts in the gene set, i.e. 30.3% of transcripts are differentially expressed.



# Gene expression at 24 hours.

## Basic QC

```{r qc_24h}
all.24h.info <- filter(sample.info, time == 24)
all.24h.sleuth <- sleuth_prep(all.24h.info)
plot_pca(all.24h.sleuth, color_by = 'treatment', text_labels = T)
```

Nothing too crazy-looking here, let's proceed.

## Esvenvalerate


Set up sample info

```{r e24.info}
e24.info <- filter(sample.info, 
                   treatment %in% c("Control", "Esfenvalerate"), 
                   time == 24)
e24.info$treatment <- factor(e24.info$treatment, levels = c("Control", "Esfenvalerate"))

e24.info
```

Set up the sleuth object

```{r e24.sleuth}
e24.sleuth <- sleuth_prep(e24.info,
                          extra_bootstrap_summary = T)
```

A quick PCA plot to check for nasties
```{r}
plot_pca(e24.sleuth, color_by = 'treatment', text_labels = T)
```

Fit the full model (with treatment effects) and the reduced model (wothout treatment effects) and do likelihood ratio tests for differential gene expression.

```{r lrt_e24}
e24.sleuth <- sleuth_fit(e24.sleuth,
                         formula = ~treatment,
                         fit_name = "full")

e24.sleuth <- sleuth_fit(e24.sleuth,
                         formula = ~1,
                         fit_name = "reduced")

e24.sleuth <- sleuth_lrt(e24.sleuth, "reduced", "full")

models(e24.sleuth)
tests(e24.sleuth)
```
Extract the LRT test results

```{r e24.results}

e24.results <- sleuth_results(e24.sleuth, test = 'reduced:full', test_type = 'lrt', show_all = FALSE)
e24.results

```


Get the results that are significant at FDR <= 0.1

```{r e24.sig.results}
e24.sig.results <- filter(e24.results,
                          qval <= 0.1)


```

How many sig results do we have?


```{r}
length(e24.sig.results$target_id)
```

Since we only have 8 significantly differentially expressed genes, we can take a look at all 8.

```{r}
x <- e24.sig.results$target_id
plot_bootstrap(e24.sleuth, x[1])
plot_bootstrap(e24.sleuth, x[2])
plot_bootstrap(e24.sleuth, x[3])
plot_bootstrap(e24.sleuth, x[4])
plot_bootstrap(e24.sleuth, x[5])
plot_bootstrap(e24.sleuth, x[6])
plot_bootstrap(e24.sleuth, x[7])
plot_bootstrap(e24.sleuth, x[8])
```

So there are 5 genes downregulated with esfenvalerate, 3 upregulated.

Lets see if there are any entries in the blast2go annoations for these genes

First, lets get the b2g annotations, such as they are



So, just for added fun, we have different identifiers -RA, presumanbly for RNA, -PA, presumably for proteins - dplyr to the rescue!

```{r e24.sig.b2g}
e24.sig.b2g <- left_join(mutate(e24.sig.results, 
                                   protein_id = str_replace(target_id, '-RA$', '-PA')), 
                            b2g.annotations)[, c(1, 15, 16)]
e24.sig.b2g

```
Nothing, but then again only 10171 (about 24%) of the transcripts have anything an entry in the blast2go annotations

Let's see if interpro scan results shed any light

Note that the downloaded file of IPS results does not include a header line. Header infor comes from ![https://github.com/ebi-pf-team/interproscan/wiki/OutputFormats]



```{r e24.sig.ips}
e24.sig.ips <- left_join(mutate(e24.sig.results, 
                                   protein_id = str_replace(target_id, '-RA$', '-PA')),
                         ips.annotations)[,c(1, 17, 18, 19, 25, 26)]
e24.sig.ips
```

Looks like AG6016473 might be an RNase.

## Thiamethoxam

Set up sample info

```{r t24.info}
t24.info <- filter(sample.info, 
                   treatment %in% c("Control", "Thiamethoxam"), 
                   time == 24)
t24.info$treatment <- factor(t24.info$treatment, levels = c("Control", "Thiamethoxam"))

t24.info
```

Set up the sleuth object

```{r t24.sleuth}
t24.sleuth <- sleuth_prep(t24.info,
                          extra_bootstrap_summary = T)
```

A quick PCA plot to check for nasties
```{r}
plot_pca(t24.sleuth, color_by = 'treatment', text_labels = T)
```

This looks OK.

Fit the full model (with treatment effects) and the reduced model (wothout treatment effects) and do likelihood ratio tests for differential gene expression.

```{r lrt_t24}
t24.sleuth <- sleuth_fit(t24.sleuth,
                         formula = ~treatment,
                         fit_name = "full")

t24.sleuth <- sleuth_fit(t24.sleuth,
                         formula = ~1,
                         fit_name = "reduced")

t24.sleuth <- sleuth_lrt(t24.sleuth, "reduced", "full")

models(t24.sleuth)
tests(t24.sleuth)
```

Extract the LRT test results

```{r t24.results}

t24.results <- sleuth_results(t24.sleuth, test = 'reduced:full', test_type = 'lrt', show_all = FALSE)
t24.results

```

Looks like nothing is significant at FDR 0.1, but lets check

```{r t24.sig.results}
t24.sig.results <- filter(t24.results,
                          qval <= 0.1)

t24.sig.results
```

Nope!

# Gene Expression at 12 Hours

## Quality control

First thing to do is understand what is going on with the outliers.

```{r}
plot_pca(all.sleuth, color_by = 'treatment', text_labels = T)
```

So, C12_rep4 and, to a lesser extend C12_rep3 are outliers on PC1. Lets take a look at the genes that are driving that.

```{r}
plot_loadings(all.sleuth, pc_input = 1)
```

We have a couple of genes that appear to be driving things: AG6039732-RA and AG6039740-RA

Let's take a look in detail at these

```{r}
plot_bootstrap(all.sleuth,
               'AG6039732-RA')

plot_bootstrap(all.sleuth,
               'AG6039740-RA')
```

Unsurprisingly, the expression levels for these genes look a bit odd for C12_rep4 and C12_rep3

```{r}
plot_bootstrap(all.sleuth,
               'AG6003545-RA')

plot_bootstrap(all.sleuth,
               'AG6008757-RA')
```
These look less "out there".

Do we have any annotation for the outlier genes?

```{r}
filter(b2g.annotations,
       protein_id %in% c('AG6039732-RA', 'AG6039740-RA'))

filter(ips.annotations,
       protein_id %in% c('AG6039732-RA', 'AG6039740-RA'))
```
Nothing!

What about the PC 2 that is separating T12_rep4?

```{r}
plot_loadings(all.sleuth, pc_input = 2)
```

AG6003545-RA already showed up in PC1.

What happens if we just get rid of AG6039732-RA and AG6039740-RA? There does not appear to be a simple way to do this in sleuth. Luckily kallisto is fast so we can re-run it using a modified transcriptome with AG6039732-RA and AG6039740-RA removed.

```{r mod.sample.info}
mod.sample.info <- mutate(sample.info, 
                          path = str_replace(sample.info$path, 'ogs', 'mod_ogs'))
```

```{r mod.all.sleuth}
mod.all.sleuth <- sleuth_prep(mod.sample.info,
                          extra_bootstrap_summary = T)

plot_pca(mod.all.sleuth, color_by = 'treatment', text_labels = T)
```

This is certainly an improvement, although C12_rep4 and C12_rep3 are still separated off on PC1. Let's take a look at the loading ...

```{r}
plot_loadings(mod.all.sleuth, pc_input = 1)
```

How do the expression levels look for the top 2 genes ...

```{r}
plot_bootstrap(mod.all.sleuth,
               'AG6040592-RA')

plot_bootstrap(mod.all.sleuth,
               'AG6003545-RA')
```

AG6040592-RA looks like a bit of an oddball, but AG6003545-RA seems OK.

Any annotations for AG6040592-RA?

```{r}
filter(b2g.annotations,
       protein_id == 'AG6040592-RA')

filter(ips.annotations,
       protein_id == 'AG6040592-RA')
```

Nope! Lets get rid of that too.

```{r mod2.sample.info}
mod2.sample.info <- mutate(sample.info, 
                          path = str_replace(sample.info$path, 'ogs', 'mod2_ogs'))
```

```{r mod2.all.sleuth}
mod2.all.sleuth <- sleuth_prep(mod2.sample.info,
                          extra_bootstrap_summary = T)

plot_pca(mod2.all.sleuth, color_by = 'treatment', text_labels = T)
```

That looks a lot better, no wild outliers anymore.

To avoid cryptic naming, copied the "mod2" kallisto output to kallisto/reduced_ogs.

```{r reduced.ogs.sample.info}
reduced.ogs.sample.info <- mutate(sample.info, 
                          path = str_replace(sample.info$path, 'ogs', 'reduced_ogs'))
```

## Esfenvalerate, 12h

Differential gene expression between control and esfenvalerate at 12h.

Set up the sample info.


```{r e12.info}
e12.info <- filter(reduced.ogs.sample.info,
            treatment %in% c("Control", "Esfenvalerate"),
            time == 12)
e12.info$treatment <- factor(e12.info$treatment, levels = c("Control", "Esfenvalerate"))
e12.info
```


Set up the sleuth object

```{r e12.sleuth}
e12.sleuth <- sleuth_prep(e12.info,
                          extra_bootstrap_summary = T)
```

A quick PCA to check nothing too strange is going on.

```{r}
plot_pca(e12.sleuth, color_by = 'treatment', text_labels = T)
```

Fit models and run liklihood ratio test.

```{r lrt_e12}
e12.sleuth <- sleuth_fit(e12.sleuth,
                         formula = ~treatment,
                         fit_name = "full")

e12.sleuth <- sleuth_fit(e12.sleuth,
                         formula = ~1,
                         fit_name = "reduced")

e12.sleuth <- sleuth_lrt(e12.sleuth, "reduced", "full")

models(e12.sleuth)
tests(e12.sleuth)
```


Get the results of the LRT

```{r e12.results}
e12.results <- sleuth_results(e12.sleuth, test = 'reduced:full', test_type = 'lrt', show_all = FALSE)
e12.results
```

Get the results that are significant at FDR = 0.1

```{r e12.sig.results}
e12.sig.results <- filter(e12.results,
                          qval <= 0.1)
e12.sig.results
```

How many genes are significantly differentially expressed?

```{r}
length(e12.sig.results$target_id)

```

Ouch!!!

This may be so many genes that is not useful, if we are going to make any sense of it, we will probably have to confine our attention to the genes with the largest expression differences. Sleuth's LRT does not yeild a measure of the effect size. However the Wald test does yeild beta, which is analgous to ln(fold_change).

On the other hand, the Wald test apparently does not perform well with small numbers of replicates. This [blog post](http://achri.blogspot.com.es/2016/10/using-kallisto-sleuth.html) suggests running the LRT for significance testing, and running the Wald test to get effect sizes.

```{r}
e12.sleuth <- sleuth_wt(e12.sleuth, 'treatmentEsfenvalerate')
e12.wald.results <- sleuth_results(e12.sleuth,
                                   test = 'treatmentEsfenvalerate')
```

Let's add the values of b to the LRT significant results.

```{r}
e12.sig.results <- left_join(e12.sig.results, 
                             e12.wald.results[,c(1,4)])

```
Get the top 25 transcripts that are significantly upregulated under esfenvalerate exposure.

```{r e12.top25}
e12.top25 <- head(arrange(e12.sig.results,
                          desc(b)),
                  n = 25)

?seq_along

```

Get the annotations for the top 25 upregulated transcripts

```{r e12.top25.b2g}
e12.top25.b2g <- filter(b2g.annotations,
                        protein_id %in% str_replace(e12.top25$target_id,
                                                    "-R",
                                                    "-P"))

e12.top25.b2g 
```

```{r e12.top25.ips}
e12.top25.ips <- filter(ips.annotations,
                        protein_id %in% str_replace(e12.top25$target_id,
                                                    "-R",
                                                    "-P"))

e12.top25.ips
```

Lets take a look at the plots of transcript counts.

```{r, warning=F}
for (id in e12.top25$target_id){
  plt <- plot_bootstrap(e12.sleuth,
                        id)
  print(plt)
  }

```


## Are the OGS trancsripts and good.

After playing around with the top 25 DEGs, I am starting to suspect there is something amiss with the OGS trancsripts. An NCBI blastx search using AG6012408-RA as the query give a couple of short hits to XP_022178029, a Myzus persicae cuticle protein. Taking XP_022178029 as a tblastn query against the SBA genome reveals hits to the ends of AG6012408-RA but also exons of several other transcript models that sit within or overlap with AG6012408-RA, some of which are on the opposite strand. This is all very fishy. Chatting with JF, he agreed it looked suspect and suggested doing my own annotation with BRAKER for comparison.



